---
title: "Peak-calling statistics exercise"
author: "Jacques van Helden"
date: "2/1/2017"
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = "", fig.width = 7, fig.height = 5)
```

# Title 1

## Title 2

### Title 3

Hola me llamo Jacques. 

*Italicos*

**bold or Bold**

***bold and italics***^[This is a footnote]


<!-- Mi secreto -->

# Probabilities

## Poisson density

$$P(X = x) = \frac{\lambda e^{-\lambda}}{k!}$$

```{r poisson_density, fig.width=10, fig.height=5, fig.cap="**Poisson density**, with $\\lambda = 3$. "}
x <- 0:20
#print(x)

p <- dpois(x, lambda=3)

plot(x, p, type="h", lwd=3,
     main="Poisson density",
     col="blue",
     xlab="X", ylab="P(X = x)")

```


# Data Loading

```{r parameters}
## Define the base URL of the course and store it in a variable
url.course <- "http://jvanheld.github.io/stats_avec_RStudio_EBA/"

## Define the fill path of the data directory, by concatenating the course URL and all the successive folders
url.data <- file.path(url.course, "practicals", "02_peak-calling", "data")
```


```{r load_chip_coverage}
## Define URL of the ChIP file
chip.bedg.file <- file.path(url.data, "FNR_200bp.bedg")
#print(chip.bedg.file)

## Load the file content in an R data.frame
chip.bedg <- read.table(chip.bedg.file)

## Set column names
names(chip.bedg) <- c("chrom", "start", "end","counts")

```

We loaded the ChIP-seq coverage profile from a bedgraph file^[`r chip.bedg.file`]. The file contained  `r nrow(chip.bedg)` lines corresponding to genomic bins. 


```{r load_genomic_input}
## Define URL of the input file
input.bedg.file <- file.path(url.data, "input_200bp.bedg")

## Load the file content in an R data.frame
input.bedg <- read.table(input.bedg.file)

## Set column names
names(input.bedg) <- c("chrom", "start", "end","counts")
```

The background was estimated with a genomic input^[`r input.bedg.file`] containing `r nrow(input.bedg)` lines.

********************************
# Results

## Distribution of read  counts per bin

```{r count distribution, fig.width=10, fig.height=8, fig.cap="**Coverage profiles**."}

par(mfrow=c(2,1))
plot(x = chip.bedg$start/1000,
     y = chip.bedg$counts,
     main="ChIP-seq coverage",
     xlab="Bin start position (Kb)",
     #     cex=0.5,
     col="blue",
     type="h",
     las=1,
     ylab="Read counts per bin")

plot(x = input.bedg$start/1000,
     y = input.bedg$counts,
     main="Genomic input coverage",
     xlab="Bin start position (Kb)",
     #     cex=0.5,
     col="grey",
     type="h",
     las=1,
     ylab="Read counts per bin")
par(mfrow=c(1,1))

```

### Interpretation of the coverage profiles

- The coverage profile of the ChIP-ped sample shows striking peaks at different genomic positions. 

- The genomic input shows fluctuations, but no such peaks. 

- Note that the scales are different: genomic input varies from `r min(input.bedg$counts)` to `r  max(input.bedg$counts)`, whereas the ChIP-seq peaks reach `r max(chip.bedg$counts)`.


