---
title: "Discovering a motif discovering appproach"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../../slides.css
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: LCG_BEII 2019
font-family: Garamond
transition: linear
---

```{r include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)
library(kableExtra)
# library(formattable)

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/R_intro_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, eval = TRUE, 
  warning = FALSE, message = FALSE, 
  results = TRUE, comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```

## Goal of the exercise

The goal of this exercise is to get an intuition of a motif discovery approach relying on the detection of over-represented oligonucleotides. 

Our approach will be pragmatic. 

We retrieved the upstream non-coding sequences of the genes involved in methionine biosynthesis and sulfur assimilation, and counted the occurrences of each hexanucleotide. 

We also computed 

- the relative frequencies (occurrences of each oligo / sum of all oligo occurrences) in the sequence of interest (the promoters of methionine-associated genes)
- the relative frequencies of ach hexanucleotide in the whole set of yeast promoters.

We would like to know if some  6nt are over-reprsented in promoters of methionine-associated genes relative to the occurrences that would be expected from a random selection of yeast promoters.

## Create a workspace for this practical

- In your home directory, create a work directory for this practical (for example `~/LCG_BEII/practical_motif_discovery/`).

```{r}
workdir <- "~/LCG_BEII/practical_motif_discovery"
dir.create(workdir, showWarnings = FALSE, recursive = TRUE)
setwd(workdir)
```


## Loading the data table


1. Download the oligonucleotide count table. 
[Scerevisiae_MET-genes_oligos-6nt-2str-noov_occ-freq.tsv](http://jvanheld.github.io/LCG_BEII/practicals/motif_discovery/data/Scerevisiae_MET-genes_oligos-6nt-2str-noov_occ-freq.tsv)

```{r download_oligo_counts}
oligo.url <- "http://jvanheld.github.io/LCG_BEII/practicals/motif_discovery/data/Scerevisiae_MET-genes_oligos-6nt-2str-noov_occ-freq.tsv"
oligo.file <- basename(oligo.url) ## Suppress the URL path and keep only the file name for local storage
download.file(oligo.url, destfile = oligo.file)
```


2. In **R**, open a new script or R markdown file. 

3. Load the data table, print the 5 top rows and the 5 bottom rows.

```{r load_oligos}

oligo.table <- read.delim(oligo.file, header = 1, row.names = 1)
# View(oligo.table)

head(oligo.table, n = 5)
tail(oligo.table, n = 5)
```

## Exploring observed and expected counts

4. Draw an histogram of the observed occurrences and evaluate the spread of counts. 

```{r obs_hist, out.width="80%", fig.width=7, fig.height=4}
x <- oligo.table$occ
range(x)
max.x <- max(x)

hist(x, breaks = 0:max.x, col = "palegreen",
     xlab = "Observed occurrences",
     ylab = "Nb of oligos",
     las = 1,
     main = "Distribution of oligonucelotide occurrences")
```


5. Draw a scatter plot comparing the observed and expected occurrences for each hexanucleotide.

```{r obs_vs_exp, out.width="80%", fig.width=7, fig.height=4.5, fig.cap="**Scatter plot of observed versus expected occurrences.** The black diagonal corresponds to the null hypothesis, the brown line denotes an arbitrary threshold on fold-change > 2. "}
exp.occ <- oligo.table$exp_occ
  
plot(exp.occ, x, col = "grey", las = 1, 
     xlab = "Expected occurrences", 
     ylab = "Observed occurrences",
     main = "Observed vs expected occurrences")
grid()
abline(a = 0, b = 1, col = "black")
abline(h = 0, col = "black")
abline(v = 0, col = "black")

abline(a = 0, b = 2, col = "brown")

```


6. Compute the ratio of observed / expected occurrences, and draw a scatter plot with this ratio (Y) as a function of the expected occurrences (X). 

```{r ratio, out.width="80%", fig.width=7, fig.height=4.5, fig.cap="**Scatter plot of observed versus expected occurrences.** The black diagonal corresponds to the null hypothesis, the brown line denotes an arbitrary threshold on fold-change > 2. "}
ratio <- (x/exp.occ)
  
plot(exp.occ, ratio, 
     col = "grey", las = 1, 
     xlab = "Expected occurrences", 
     ylab = "(obs/exp) ratio",
     main = "(obs/exp) ratio")
grid()
abline(h = 1, col = "black")
abline(h = 2, col = "brown")

```


6. Compute the log-ratio of observed / expected occurrences, and draw a scatter plot with this log-ratio (Y) as a function of the expected occurrences (X). 

```{r lr, out.width="80%", fig.width=7, fig.height=4.5, fig.cap="**Scatter plot of observed versus expected occurrences.** The black diagonal corresponds to the null hypothesis, the brown line denotes an arbitrary threshold on fold-change > 2. "}
lr <- log(x/exp.occ)
  
plot(exp.occ, lr, 
     col = "grey", las = 1, 
     xlab = "Expected occurrences", 
     ylab = "log(obs/exp)",
     main = "Log-ratio")
grid()
abline(h = 0, col = "black")
abline(h =  log(2), col = "brown")

```


$$lr = log(x/<X>)$$

7. Compute the log-likelihood ratio ($llr$), defined below, and draw a scatter plot with this $llr$ as a function of the expected occurrences. 

$$llr = f \cdot log(x/<X>)$$


```{r llr, out.width="80%", fig.width=7, fig.height=4.5, fig.cap="**Scatter plot of log-likelihood ratio (llr) versus expected occurrences.** The black line corresponds to the null hypothesis, the brown line denotes an arbitrary threshold on fold-change > 2. "}
p <- oligo.table$exp_freq
llr <- p * log(x/exp.occ)
  
plot(exp.occ, llr, 
     col = "grey", las = 1, 
     xlab = "Expected occurrences", 
     ylab = "llr",
     main = "Log-likelihood ratio")
grid()
abline(h = 0, col = "black")
# abline(h =  log(2), col = "brown")

```



## Computing over-representation significance

8. Draw a binomial distribution with parameters $n = 8000$, $p = 0.0001$.

```{r}
n <- 8000
p <- 0.001
x <- 15# Number of successes
X <- 0:40 ## values to display

plot(X, dbinom(x = X, size = n, prob = p), type = "h")
arrows(x, 0.04, x, 0.02, lwd = 2, length = 0.1, angle = 30, col = "red")


pval <- pbinom(q = x - 1, size = n, prob = p, lower.tail = FALSE)
legend("topright", legend = paste("pval =", signif(digits = 3, pval)))

```


8. Use the binomial distribution to compute the P-value of the observed occurrences. 

$$P = T(X \ge x)$$

```{r}
x <- oligo.table$obs_freq ## Nuumber of successes
n <- sum(x)   ## Number of trials
p <- oligo.table$exp_freq ## Success probability


```


9. Draw an histogram with the P-values of all hexanucleotides, with 20 bins.

10. Draw a scatter plot with the P-value (Y) as a function of the log-ratio (X). 

11. Compute the E-value, and the significance. 

$$E = P \cdot N$$
$$sig = -log_{10}(E)$$

12. Draw a **Volcano plot**, with the significance as a function of the log-ratio. 

13. Compute the P-value using the Poisson distribution as approximation of the binomial. Are we in suitable conditions for this approximation ? Draw a plot comparing the P-values obtained by the binomial and Poisson distributions. 

14. Compute the P-value using a normal approximation of the binomial distribution. 

    a. Are we in suitable conditions to approximate a binomial with a normal ?
    b. Compare the P-values obtained with the binomial and norma distrinbutions, resp. 
    





