---
title: "Practical: discovering a motif discovery algorithm"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    slide_level: 2
    self_contained: no
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  revealjs::revealjs_presentation:
    theme: night
    transition: none
    self_contained: true
    css: ../../slides.css
  slidy_presentation:
    smart: no
    slide_level: 2
    self_contained: yes
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle: LCG_BEII 2019
font-family: Garamond
transition: linear
---

```{r include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)
library(kableExtra)
# library(formattable)

options(width = 300)
# options(encoding = 'UTF-8')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/R_intro_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, eval = TRUE, 
  warning = FALSE, message = FALSE, 
  results = FALSE, comment = "")

options(scipen = 12) ## Max number of digits for non-scientific notation
# knitr::asis_output("\\footnotesize")

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this tutorial, we will use **R** to detect over-represented motifs in the regulatory sequences of a set of genes involved in a biological process of interest. 

**Our study case:** We examplify the approach with the **methionine biosynthetic process** in the yeast *Saccharomyces cerevisiae*. 

Each student will be invited to run the same analyses with a different biological process, and we will compare the results gathered by all of us in order to draw some general insight into the relevance of the approach.

## Suggested biological processes

I provide here some examples of biological processes of the yeast Saccharomyces cerevisiae. Note that these processes may have different names depending on the databases, so you will have to adapt them in different steps of the tutorial. 

- L-methionine biosynthesis
- L-arginine biosyntesis
- L-lysine biosynthesis
- any other aminoacid metabolic pathway
- Ergosterol biosynthesis
- Galactose utilization
- Phosphae utilization
- Nitrogen utilization
- ... (don't hesitate to be adventurous)

## Resources

| Resource | URL |
|-----------------------------|--------------------------------------------|
| Gene Ontology Consortium | <http://www.geneontology.org/> |
| Ensembl BioMart | <https://www.ensembl.org/> |
| RSAT | Regulatory Sequence Analysis Tools (<http://rsat.eu/>) |


## Viewing metabolic pathways on BioCyc

- Open a connection to **BioCyc** (<https://biocyc.org/>)
- In the right-left corner, click **change organism database** and select *Saccharomyces cerevisiae S288c*.
- Select the function **Metabolism** > **Generate a metabolic map poster** to get an overview of the whole metabolism.
- Select the tool *Browse pathway ontology* and select a pathway of your choice (not too big, not too small, let us say between 5 and 50 genes). 

- As an example I select two pathways: 
    - "superpathway of sulfur amino acid biosynthesis"
    - "L-methionine de novo biosynthesis" 
    
- Click on the pathway instances to display the pathway maps. PLay with the options to increase/decrease the level of detail. 

- In your custom result folder, save a figure that provides a nice view of the pathway, and could be used for example in the introduction of a report. 


## Finding a process in the Gene Ontology

- Open a connection to the Gene Ontology Consortium (<http://www.geneontology.org/>)

- In the menu at the top of the page, select **Tools** > **AmiGO 2**. 

- In the quick search box, type the name of your biological process of interest (e.g. 'methionine biosynthetic process'). As you type, a pop-up menu appears to display the subset of GO terms corresponding to your query. Select the one you need (e.g. "L-methionine biosynthetic process").

This opens the record describing the biological process you chose. 

- Before anything else, take note of the GO dientifier (e.g.: GO:0071265)

- Read the **Term description** box and check that the description corresponds to your choice. 

- At the bottom of the **Term description** page, click on the **Link to all genes and gene products annotated to [your] process**.

- In the filter box, type "Saccharomyces cerevisiae" and count the number of genes associated to your process in the budding yeast. 

## Getting genes from BioMart

- Open a connection to **Ensembl** (<https://www.ensembl.org/>) and select  the **BioMart** tool. 

- **Choose the database:** *Ensembl Genes [version]* (note: the current version is 95 in Feb 2019)

- **Choose dataset:** *Saccharomyces cerevisiae genes (R-46-1-1)*

- In the left menu, click **Filters** and open the section **Gene Ontology**.

- Type the GO term accession ID of your choice (e.g. GO:0071265).

- In the top corner of the windfow, click **Count** and check that you have a reasonable number of genes (with GO:0071265, there are 13 genes). 

We will now customize the attributes to be extracted from BioMart.

- In the left menu, click **Attributes**, then **Genes**

- Activate the following attributes: 

    - Gene stable ID
    - Protein stable ID
    - Gene description
    - Gene name

Note: BioMart is a flexible way to quickly get a lot of information about genes. For other tasks you might be led to select additional attributes. 

- In the top left corner, click **Results** to get the selected genes.

- Select **Export all results to File** and **TSV** (tab-separted values) and click **Go**.  This will download a tab-delimited text file in tge Download folder of your Web browser. Save this file on your computer. 


- Rename the downloaded file (e.g. `Scerevisiae_methionine-biosynthetic-process_GO-0071265_genes.tsv`).

- Open the TSV file to check its content. The most convenient is to open it with a spreadsheet tool (e.g. LibreOffice Calc, Excel), but you can also view it with a simple text editor. 


At this stage, we dispose of a table with the description of all the genes associated to the biological process of our choice.



## Discovering motifs in promoters of functionally related genes

We will try to discover over-represented motifs in the promoters of the genes involved in our pathway of interest. 

- Open the TSV file with the genes involved in your biological process. 
- Select the colum with stable gene identifiers. In my case I have the 13 following IDs. 


```
YAL012W
YPL273W
YJR148W
YHR137W
YJR024C
YJR024C
YEL038W
YEL038W
YMR009W
YMR009W
YHR112C
YGL202W
YGL184C
YGL184C
YHR208W
YPR118W
YLR017W
```

- Open a connection to the fungal RSATv server (<http://fungi.rsat.eu/>)
- With the tool **retrieve sequence**, get the non-coding sequences located upstream of your genes. 

- In the **Next Step** box of the rsult page, click the button **oligo-analysis**. 

- For **oligomer lengths** select *6* (uncheck the values 7 and 8 that are selected by default). 

- Leave all other options unchanged and click **GO**








